Hydras, Ordinals \& Co. \\ A library in Coq of entertaining formal mathematics

### 2021-10-14 doc/JFLA2022/paper.tex
another pass over comparison

The Gaia project by Jos\'e Grimm aimed to formalize mathematics in the style of N. Bourbaki in Coq. The formalization of the first book by Bourbaki, on the theory of sets, was initially described in a technical report in July 2009~\cite{Grimm2009a}. Grimm then wrote (and continually updated) technical reports describing the formalization of Bourbaki's two subsequent books~\cite{Grimm2009b,Grimm2016} and additional topics in number theory~\cite{grimm:hal-00911710,Grimm2014}, before he passed away in 2019.

@@

We believe that Coq books can broadly be divided into two categories:
\begin{enumerate}
\item[a)] \emph{Executable textbooks} that are in effect large, well-commented Coq programs from which other representations of the material (HTML, PDF) are derived using tools.
\item[b)] \emph{Traditional textbooks} generated from documentation languages such as LaTeX that are accompanied by standalone Coq formalizations and/or code snippets.
\end{enumerate}

Table~\ref{tbl:books} shows a categorization of the Coq books in English we are aware of.

@@

We believe the \Hydras electronic book finds good tradeoffs between the advantages of executable books and traditional books.

>>>>

The Gaia project by Jos\'e Grimm aimed to formalize mathematics in Coq in the style of Nicolas Bourbaki. The formalization of the first book in the Elements of Mathematics series by Bourbaki, on the theory of sets, was initially described in a technical report in July 2009~\cite{Grimm2009a}. Grimm then wrote (and continually updated) technical reports describing the formalization of Bourbaki's two subsequent books~\cite{Grimm2009b,Grimm2016} and additional topics in number theory~\cite{grimm:hal-00911710,Grimm2014}, before he passed away in 2019.

@@

We believe that Coq books can broadly be divided into two categories:
\begin{itemize}
\item \emph{Executable textbooks} that are in effect large, well-commented Coq programs from which other representations of the material (HTML, PDF) are derived using tools.
\item \emph{Traditional textbooks} generated from documentation languages such as LaTeX that are accompanied by standalone Coq formalizations and/or code snippets.
\end{itemize}
Table~\ref{tbl:books} shows a category breakdown for what we believe are the most prominent Coq books in English.

@@

We believe the \Hydras electronic book finds good tradeoffs between the properties of executable books and traditional books. Specifically, in an executable book, the proof assistant language places limits on the structure. For example, Coq may not permit repeating a definition from another module verbatim, since the definition uses section variables that are not present in the current context. Thanks to our \alectr based toolchain, the \Hydras book can include ``live'' code fragments at any point in the text. And in contrast to traditional textbooks, we validate the consistency of code fragments during every build of the electronic book. We follow executable books in providing ``continuous delivery'' of new revisions at a high pace, but we also aim to make regular timestamped versions compatible with specific versions of Coq.
